---
title: 'Human Activity Recognition: Weight Lifting Exercises Dataset Analysis'
output: pdf_document
---

```{r echo=FALSE, eval=FALSE}
if(FALSE) {

project-PML.Rmd (Practical Machine Learning project)

About the data set:

This data set attempts to define quality of execution and investigate three aspects that pertain to qualitative activity recognition: the problem of specifying correct execution, the automatic and robust detection of execution mistakes, and how to provide feedback on the quality of execution to the user. This data set utilizes the on-body sensing approach.

Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate. The exercises were performed by six male participants aged between 20-28 years, with little weight lifting experience. We made sure that all participants could easily simulate the mistakes in a safe and controlled manner by using a relatively light dumbbell (1.25kg).

The primary code to preprocess the data is located [here](http://github.com/bpoweski/practical-machine-learning/blob/master/project.R).

There is additional code within the knitr markdown file [here](http://github.com/bpoweski/practical-machine-learning/blob/master/project.Rmd).

#### The goal of your project is to predict the manner in which they did the exercise. 
This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online.
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 

}
```

```{r setup, echo=FALSE}
rm(list=ls())
options(digits=4, width=100)

if (file.exists("c:/windows/win.ini")) {
  setwd("w:/mooc/Practical Machine Learning/project") 
  } else setwd("/home/derec/mnt/project/mooc/Practical Machine Learning/project") 
```


#### Data Processing

```{r load_libraries, echo=TRUE}
# load necessary libraries
suppressWarnings(suppressMessages(library(caret)))
suppressWarnings(suppressMessages(library(corrplot)))
suppressWarnings(suppressMessages(library(FactoMineR)))
suppressWarnings(suppressMessages(library(grDevices)))
suppressWarnings(suppressMessages(library(knitr)))
suppressWarnings(suppressMessages(library(randomForest)))
```


```{r load_data}
# download files if they do not exist
if (!file.exists("pml-training.csv")) {
    download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", 
        destfile = "pml-training.csv")
}

if (!file.exists("pml-testing.csv")) {
    download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", 
        destfile = "pml-testing.csv")
}

# read training set and handle na strings
tmp.train <- read.csv("pml-training.csv", na.strings=c("NA","","#DIV/0!"," "), stringsAsFactors=T)
names(tmp.train)[1] <- "id"

# do not keep columns with mostly missing values or irrelevant fields.
keep <- c("id","roll_belt", "pitch_belt", "yaw_belt", "total_accel_belt", 
    "gyros_belt_x", "gyros_belt_y", "gyros_belt_z", "accel_belt_x", "accel_belt_y", 
    "accel_belt_z", "magnet_belt_x", "magnet_belt_y", "magnet_belt_z", "roll_arm", 
    "pitch_arm", "yaw_arm", "total_accel_arm", "gyros_arm_x", "gyros_arm_y", 
    "gyros_arm_z", "accel_arm_x", "accel_arm_y", "accel_arm_z", "magnet_arm_x", 
    "magnet_arm_y", "magnet_arm_z", "roll_dumbbell", "pitch_dumbbell", "yaw_dumbbell", 
    "total_accel_dumbbell", "gyros_dumbbell_x", "gyros_dumbbell_y", "gyros_dumbbell_z", 
    "accel_dumbbell_x", "accel_dumbbell_y", "accel_dumbbell_z", "magnet_dumbbell_x", 
    "magnet_dumbbell_y", "magnet_dumbbell_z", "roll_forearm", "pitch_forearm", 
    "yaw_forearm", "total_accel_forearm", "gyros_forearm_x", "gyros_forearm_y", 
    "gyros_forearm_z", "accel_forearm_x", "accel_forearm_y", "accel_forearm_z", 
    "magnet_forearm_x", "magnet_forearm_y", "magnet_forearm_z","classe")

train <- tmp.train[,keep]

# load test set and conform it to match training set.
tmp.test <- read.csv("pml-testing.csv", na.strings=c("NA","","#DIV/0!"," "), stringsAsFactors=T)
names(tmp.test)[1] <- "id"
tmp.test$id <- tmp.test$problem_id # copy problem_id to row number field and use as id
tmp.test <- tmp.test[, -160]
tmp.test$classe <- ""
tmp.test <- tmp.test[,keep]
tmp.test$classe <- factor(tmp.test$classe, levels=c("A","B","C","D","E"))
test <- tmp.test

# dimensions of data sets
dim(train)
dim(test)

# remove unneeded variables.
rm(tmp.test,tmp.train)
```

#### Feature Selection

The data set for analysis has 160 attributes, so exploratory analysis with scatter plot matrices is impractical. The first step of this analysis will be the attempt to reduce the number of attributes prior to modeling. Instead of allowing the model to select which attributes to use from the raw data set, I will filter the data to remove irrelevant, and highly correlated attributes prior to modeling.


```{r data_reduction}
vis.train <- train[,2:ncol(train)-1]

# scale and center training set without id or classe attributes.
vis.train.scale <- scale(vis.train, center=TRUE, scale=TRUE)
corMat <- cor(vis.train.scale) #compute the correlation matrix
```

Plot 1: Correlation plot of initial set of predictors. Extreme correlation values are positioned at the left and right edges.

```{r plot_1, echo=FALSE, fig.height=10, fig.width=10}
corrplot(corMat, order="FPC", method="color", type="lower", tl.cex=0.75, tl.col=rgb(0,0,0),
         title="")
```


```{r}
# remove highly correlated attributes and re-plot.
highlyCor <- findCorrelation(corMat, 0.80) # set highly correlated at 0.80
vis.Filtered.scale <- vis.train.scale[,-highlyCor]
corMat <- cor(vis.Filtered.scale)
```

Plot 2: Correlation plot of revised set of predictors.

```{r plot_2, echo=FALSE, fig.height=10, fig.width=10}
corrplot(corMat, order="FPC", method="color", type="lower", tl.cex=0.8, tl.col=rgb(0, 0, 0), title="")

vis.keep <- colnames(vis.Filtered.scale)
vis.keep[42] <- "classe"

train <- train[,vis.keep]
test <- test[,vis.keep]

# remove unneeded variables.
rm(corMat,vis.Filtered.scale,vis.train,vis.train.scale,highlyCor,vis.keep)
```


#### Modeling

Having removed problem attributes, and those that are highly correlated with each other, I will fit a clustering model to predict the outcomes of the testing set.

```{r randomForest, cache=TRUE}
set.seed(123)
mod.train <- train[, 2:ncol(train)]
mod.test <- test[, 2:ncol(train)]
model <- randomForest(classe ~ ., data=mod.train, ntree=100)
model
mod.test$classe <- predict(model, mod.test)
mod.test$classe
```

The predicted values proved to be 100% accurate. The predicted outcomes were nearly identical using 50 trees as they were using 2048 trees.


#### Principal Component Analysis

```{r pca, fig.height=10, fig.width=10}
tmp <- read.csv("pml-training.csv", na.strings=c("NA","","#DIV/0!"," "), stringsAsFactors=T)
names(tmp)[1] <- "id" # row number is unique, so keep as id field
tmp.train <- tmp[,keep] # do not keep columns with mostly missing values or timestamps.

# PCA with function PCA
pca.train <- tmp.train
pca <- PCA(pca.train[,2:53], scale.unit=TRUE, ncp=5, graph=F) # scale all features
#plot.PCA(pca, axes=c(1, 2), choix="ind", habillage="ind")
plot.PCA(pca, axes=c(1, 2), choix="ind", habillage="none")
#dimdesc(pca) # sort the variables most linked to each principal component

# remove unneeded variables.
rm(pca.train,tmp.train)
```


```{r submittal, echo=FALSE}
# pml_write_files = function(x){
#   n = length(x)
#   for(i in 1:n){
#     filename = paste0("submit/problem_id_", i, ".txt")
#     write.table(x[i], file=filename, quote=FALSE, row.names=FALSE, col.names=FALSE)
#   }
# }
# pml_write_files(test$classe)
```

------------------------------------------------------------------------------------

Data for this analysis courtesy of:

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013. 

References:

[1] http://www.r-bloggers.com/introduction-to-feature-selection-for-bioinformaticians-using-r-correlation-matrix-filters-pca-backward-selection/