---
title: 'Human Activity Recognition: Weight Lifting Exercises Dataset Analysis'
output: pdf_document
---

```{r echo=FALSE, eval=FALSE}
if(FALSE) {

project.Rmd (Practical Machine Learning project)

About the data set:

This data set attempts to define quality of execution and investigate three aspects that pertain to qualitative activity recognition: the problem of specifying correct execution, the automatic and robust detection of execution mistakes, and how to provide feedback on the quality of execution to the user. This data set utilizes the on-body sensing approach.

Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate. The exercises were performed by six male participants aged between 20-28 years, with little weight lifting experience. We made sure that all participants could easily simulate the mistakes in a safe and controlled manner by using a relatively light dumbbell (1.25kg).

#### The goal of your project is to predict the manner in which they did the exercise. 
This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online.
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 

}
```

```{r setup, echo=FALSE}
rm(list=ls())
options(digits=4, width=100)

if (file.exists("c:/windows/win.ini")) {
  setwd("w:/mooc/Practical Machine Learning/project") 
  } else setwd("/home/derec/mnt/project/mooc/Practical Machine Learning/project") 
```

## Data Processing


#### Load Necessary Libraries

```{r load_libraries, echo=T}
suppressWarnings(suppressMessages(library(caret)))
suppressWarnings(suppressMessages(library(grDevices)))
suppressWarnings(suppressMessages(library(knitr)))
suppressWarnings(suppressMessages(library(randomForest)))
```

#### Loading and Pre-processing

```{r load_data, cache=TRUE}
tmp <- read.csv("pml-training.csv", na.strings=c("NA","","#DIV/0!"," "), stringsAsFactors=T)
#dim(tmp)
names(tmp)[1] <- "id" # row number is unique, so use as id field
tmp.complete <- Filter(function(x)(length(unique(x))>1), tmp) # keep columns with > 1 unique value
#keep <- colnames(tmp.complete) # save names of columns kept and apply to test set
keep <- c("id","roll_belt", "pitch_belt", "yaw_belt", "total_accel_belt", 
    "gyros_belt_x", "gyros_belt_y", "gyros_belt_z", "accel_belt_x", "accel_belt_y", 
    "accel_belt_z", "magnet_belt_x", "magnet_belt_y", "magnet_belt_z", "roll_arm", 
    "pitch_arm", "yaw_arm", "total_accel_arm", "gyros_arm_x", "gyros_arm_y", 
    "gyros_arm_z", "accel_arm_x", "accel_arm_y", "accel_arm_z", "magnet_arm_x", 
    "magnet_arm_y", "magnet_arm_z", "roll_dumbbell", "pitch_dumbbell", "yaw_dumbbell", 
    "total_accel_dumbbell", "gyros_dumbbell_x", "gyros_dumbbell_y", "gyros_dumbbell_z", 
    "accel_dumbbell_x", "accel_dumbbell_y", "accel_dumbbell_z", "magnet_dumbbell_x", 
    "magnet_dumbbell_y", "magnet_dumbbell_z", "roll_forearm", "pitch_forearm", 
    "yaw_forearm", "total_accel_forearm", "gyros_forearm_x", "gyros_forearm_y", 
    "gyros_forearm_z", "accel_forearm_x", "accel_forearm_y", "accel_forearm_z", 
    "magnet_forearm_x", "magnet_forearm_y", "magnet_forearm_z","classe")
#dim(tmp.complete)
#colnames(tmp.complete)
train <- tmp.complete[,keep]
#dim(train)
#colnames(train)
#write.csv(train,"train.csv",row.names=F) # save for RM input

tmp <- read.csv("pml-testing.csv", na.strings=c("NA","","#DIV/0!"," "), stringsAsFactors=T)
tmp$X <- tmp$problem_id # copy problem_id to row number field and use as id
names(tmp)[1] <- "id" # label attribute that matches training set label name
#colnames(tmp)
tmp <- tmp[,1:159] # do not include "problem_id" field
tmp$classe <- 'Unk' # add predicted attribute and label as unknown
test <- tmp[, keep] # use columns that match training set
#dim(test)
#colnames(test)
#write.csv(test,"test.csv",row.names=F) # save for RM input
```


#### Create reduced training set by identifying and eliminating 95%+ correlated attributes.
```{r data_reduction}
#str(train)


# identify correlated predictors
#suppressMessages(library(caret))
#train.cm <- train[,7:153]
#descrCor <- cor(train.cm, use="na.or.complete")
#summary(descrCor[upper.tri(descrCor)])
#noEcho <- findCorrelation(descrCor)

# use cutoff of 95% correlation to filter
#highlyCorDescr <- findCorrelation(descrCor, cutoff = 0.95)
#filteredDescr <- train.cm[, -highlyCorDescr]
#descrCor2 <- cor(filteredDescr)
#summary(descrCor2[upper.tri(descrCor2)])

# create reduced attribute training set
#train.r <- train[,-highlyCorDescr]
```

#### Transformations

```{r }
#unique(train$new_window)
#unique(train$skewness_yaw_dumbbell)
#train.r <- train[,sapply(train, function(v) var(v, na.rm=TRUE)!=0)]
#str(train.r)
#dim(train)
```

#### Modeling

##### Decision Tree Classifier
```{r}
# library(rpart)
# options(digits=3)
# set.seed(123)
# tmp.train <- train[,2:54]
# tmp.test <- test[,2:54]
# 
# # create decision tree
# treemodel1 <- rpart(classe ~ ., data=tmp)
# printcp(treemodel1)
# 
# # predict misclassification count using the tree
# tmp.test$classe <- predict(treemodel1, newdata=tmp.test, type='class')
```


```{r randomForest, cache=TRUE}
set.seed(123)
tmp.train <- train[,2:54]
tmp.test <- test[,2:54]

model <- randomForest(classe ~ ., data=tmp.train, ntree=2048)
model
tmp.test$classe <- predict(model, tmp.test)
```


```{r submittal, echo=FALSE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("submit/problem_id_",i,".txt")
    write.table(x[i], file=filename, quote=FALSE, row.names=FALSE, col.names=FALSE)
  }
}
pml_write_files(tmp.test$classe)
```

------------------------------------------------------------------------------------

Data for this analysis courtesy of:

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013. 